Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-01-23 E-Hypergraphs-LICS24.tex
Typos, alignment

Although equality saturation is already a state-of-the-art algorithmic optimisation technique, there is an opportunity to better understand its mathematical foundations. 
We propose to address this issue via a categorical axiomatisation for e-graphs and their rewriting. 
Considering programs as represented by terms of an arbitrary algebraic theory, we demonstrate a correspondence between e-graphs and (string diagrams for) simple \textit{semilattice-enriched} Cartesian categories. 
The approach we take is guided by the established methodology of creating a correspondence between string diagrams and graph rewriting, as encountered in a large body of research. 
The starting innovation of our approach is the semilattice-enrichment. 
The \textit{join} operator lifts the formalism so that it can now operate with (non-empty) \textit{sets} of terms (string diagrams, morphisms); the rest of the paper is about working out the mathematical implications of introducing this enrichment to string diagrams and their concrete combinatorial represetation.

@@

The translation from e-graphs to enriched string diagrams is illustrated informally below,
noting how the typing constraints of the $+$ constructor are satisfied in the image of the translation by discarding in each component all unnecessary inputs.
\[
    \tikzfig{categorical-semantics/egraph-translation-1}
\]
Examples of this translation are given in the second row of Figure \ref{fig:e-graph-example}. In particular, note that the translation of the \textit{cyclic} e-graph (e)
requires the use of a categorical \textit{trace}, generating a cycle. In this paper, we focus on \textit{acyclic} e-graphs and their corresponding catgeories; but the above example illustrates that the extension to the cyclic case seems to be routine. As it would involve needless additional presentational complexity we leave it as further work.

@@

\subsection{Symmetric Monoidal Theories and PROPs. }

A presentation of an algebraic theory is traditionally given by a \textit{signature} of $n$-ary operations and a set of \textit{equations}---formally,  pairs of terms freely generated over the signature.  We are interested here in symmetric monoidal theories, which are generalizations of algebraic theories where the operations of the signature may have arbitrary \textit{co-arities}.  First,  we generalize the notion of a signature. 
\begin{definition}[Monoidal signature]
A \textit{(monoidal) signature} $\Sigma$ is given by a set of \textit{generators} $c: n \to m$,  with \textit{arity} $n$ and \textit{coarity} $m$.  %A \textit{signature homomorphism} $h: \Sigma \to \Gamma$ is a function from $\Sigma$ to $\Gamma$ which preserves the (co-)arities of elements. 
\end{definition}

@@

We will use enrichment to axiomatize the e-box structure as the ``join" of two morphisms $f,g: A \to B$ as $f + g: A \to B$. Note, in particular,  that semilattices do not require a unit for $+$. This is important, since, in any commutative monoid enriched category,  if there exists a categorical product, then it must be a categorical biproduct \cite{maclane}---a degeneracy we wish to avoid.  More precisely our hom-sets will be given the structure of a semilattice.  
Generally,  we may define a category \textit{enriched} in any other monoidal category $M$ (\textit{e.g.},  \textbf{SLatt}).  In this case,  hom-sets are generalized to hom-objects of $M$ (\textit{e.g.},  a semilattice) and composition 
\[
	\circ: Hom(B,C) \otimes Hom(A,B) \to Hom(A,C)
\]
is defined as a $M$-morphism.  In other words,  composition must respect the additional structure on hom-sets.  
In our case,  the category we wish to enrich is also monoidal,  so we additionally ask that the monoidal structure also respects this structure.  Technically, this amounts to asking that the tensor product is an enriched functor. 
We omit further details of the general case of enrichment here, and instead work with the following more concrete defintion, where $M = \textbf{SLatt}$.

@@

In ~\cite{bonchi_string_2022-1},  hypergraphs \textit{with interfaces} and their DPO rewriting are shown to be sound and complete for SMTs including a Frobenius algebra. 
Hypergraphs generalize graphs by allowing edges to have multiple sources and targets. 
When interpreting string diagrams as hypergraphs,  morphisms are interpreted as edges,  and wires as vertices.  
However,  there is also a need to specify which vertices are to be considered as \textit{input} and \textit{outputs},  corresponding to the dangling wires at the bottom and top of a string diagram. 
This gives rise to the definition of a hypergraph with interfaces.  
Subsequent work \cite{bonchi_string_2022-2} builds on this correspondence to give a restriction of hypergraphs with interfaces (called \textit{monogamous} directed acyclic),  leading to a correspondence between symmetric monoidal string diagrams and the category of such hypergraphs.

@@

The intuitive notion of (hyper-)graph rewriting is  as expected: given a rewrite rule $\mathcal L \rightsquigarrow \mathcal R$ and a graph $\mathcal{G}$,  we expect to find an occurrence of $\mathcal L$ within $\mathcal{G}$,  remove it,  and then plug $\mathcal R$ into the hole to recover the rewritten graph $\mathcal{H}$.  In DPO rewriting,  such a rewrite rule is formalized as a span $\mathcal L \xleftarrow{} \mathcal K \xrightarrow{} \mathcal R$ in some appropriate category of graphs,  where $\mathcal K$ is some invariant that should be maintained during rewriting.
A typical DPO square is depicted in Figure~\ref{fig:dpo_dpoi}~(left).

@@

To express the rewriting of (hyper-)graphs with interfaces,  \textit{i.e.},  the ones that represent morphisms in $\HypI{\Sigma}$, the DPO formalism has been extended to double-pushout rewriting with \textit{interfaces}  (DPOI) as shown in Figure~\ref{fig:dpo_dpoi}~(right).
In the case of $ \catname{Hyp(\Sigma)}$,  we wish rewrite rules to be pairs of discrete cospans of hypergraphs with matching interfaces: $n \xrightarrow{} \mathcal L \xleftarrow{} m$, $n \xrightarrow{} \mathcal R \xleftarrow{} m$.  
By noting that a cospan $0 \xrightarrow{} \mathcal L \xleftarrow{[f,g]} n + m$ encodes the same data as a cospan $n \xrightarrow{f} \mathcal L \xleftarrow{g} m$ (Remark 3.14~\cite{bonchi_string_2022-1}) we can formulate a rewrite rule as a single span $\mathcal L \xleftarrow{} n+m \xrightarrow{} \mathcal R$,  and we can do similarly for the discrete cospans giving the interface to $\mathcal G$ and $\mathcal H$.  
We now make this identification throughout without further mention.  
Thus we reach the following variant of DPO rewriting.
\begin{definition}[DPOI rewriting]\label{def:dpoi}
Given a span of morphisms $\mathcal G \xleftarrow{} n+m \xrightarrow{} \mathcal H$ in $\textbf{Hyp}(\Sigma)$,  we say \textit{$\mathcal G$ rewrites to $\mathcal H$ with interface $n+m$ via rewrite rule $\mathcal L \xleftarrow{} i+j \xrightarrow{} \mathcal R$} if there exists an object $\mathcal C$ and morphisms which complete the following commutative diagram such that the two marked squares are pushouts.
% \[
% \begin{tikzcd}
% \mathcal L \arrow[d, "m"'] & i+j \arrow[d] \arrow[l] \arrow[r] & \mathcal R \arrow[d]   \\
% \mathcal{G}^{\urcorner}     & \mathcal C \arrow[l] \arrow[r]                      & ^{\ulcorner}\mathcal H \\
%                   & n+m \arrow[lu] \arrow[u] \arrow[ru]          &              
% \end{tikzcd}
% \]
\begin{figure}[t!]
    \begin{subfigure}{0.45\linewidth}
    \[
            \scalebox{0.8}{
                \tikzfig{combinatorial_semantics/DPO_square}
            }
    \] 
    % \subcaption{DPO square}
    % \label{fig:dpo}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.45\linewidth}
        \[
    \scalebox{0.8}{
        \tikzfig{combinatorial_semantics/DPOI_square_Hyp}
    }
\]
    % \subcaption{DPOI square}
    % \label{fig:dpoi}
    \end{subfigure}
    \captionsetup{belowskip=-3ex}
    \caption{DPO and DPOI squares}
    \label{fig:dpo_dpoi}
\end{figure}

\end{definition}
We need to restrict the notions of pushout complement and of match in order to preserve monogamy and directed acylcicity.

@@

Hierarchical (hyper-)graphs are a well-studied area of research \cite{plump:hierarchical-graphs, montanari:gs-lambda, palacz:hierarchical-transform, Gaducci:hierarchical-graphs, Ghica:hierarchical}.  In this section,  we follow analogous steps to those used to achieve a combinatorial representation of $\textbf{PROP}(\Sigma)$ by $\mathbf{MHypI}(\Sigma)$,  but in the semilattice enriched setting.  First we define an \textit{e-hypergraph},  which supports a hierarchical notion of ``e-box'' with distinct components. Then,  we extend this notion with interfaces via a generalization of the cospan construction.  Finally,  we restrict the morphisms under consideration to those satisfying the MDA condition,  as in the previous section.  An example diagram representing e-hypergraphs with interfaces can be found in Figure \ref{fig:A+B}.

@@

When modelling enriched string diagrams using e-hypergraphs,  we require to keep track of more than simply the ultimate external inputs and outputs of the diagram: each internal e-box has its own inputs and outputs,  which we must also keep track of.  Thus,  we introduce an \textit{extended} cospan construction.  Since the $\catname{EHyp(\Sigma)}$ does not have pushouts in general,  we skip straight to considering a category of extened cospans of discrete e-hypergraphs (\textit{i.e.},  discrete hypergraphs considered as e-hypergraphs in the obvious way),  for which pushouts do exist.

We use the notation $n \setminus m$ to denote the discrete e-hypergraph with $n - m$ vertices,  in particular with the vertices of $m$ removed from $n$ when vertices of $m$ is a sub-e-hypergraph of $n$.
\begin{definition}[Category of e-hypergraphs with interfaces]
    The category of \textit{e-hypergraphs with extended interfaces} $\Ecospans$ has discrete e-hypergraphs as objects,  with homsets $\Ecospans(n,m)$ consisting of isomorphism classes of \textit{extended cospans}, defined as follows:  
    \[
    n \xrightarrow{f_{ext}} n' \xrightarrow{f_{int}} \mathcal{G} \xleftarrow{g_{int}} m' \xleftarrow{g_{ext}} m
    \]
    where $\mathcal{G}$ is an e-hypergraph,  and $n, n', m, m'$ are discrete e-hypergraphs,  $f_{ext},f_{int},g_{ext},g_{int}$ are monomorphisms in $\catname{EHyp(\Sigma)}$,  and the image of $f_{ext};f_{int}$ and of $g_{ext};g_{int}$ consist exclusively of top-level vertices,  and that that vertices in the strictly internal interface (defined below) are not top-level.  We will sometimes write $\mathcal G$ to denote the extended cospan,  where it is clear from context $\mathcal G$ is equipped with extended interfaces.

@@

Unlike in the previous section,  in order to give an equivalence with $\textbf{PROP}^+(\Sigma)$,  we must first develop the theory of DPOI rewriting for e-hypergraphs.  This is beacuse the equations for semilattice enrichment are not intended to be subsumed by the e-hypergraph representation.  Instead,  we will treat them via rewriting,  as we do with equations of the signature.  This is because the semilattice equations involve sharing and copying, operations on string diagrams which are not usually quotiented away, cf. string diagrams for Cartesian categories.  An ideal combinatorial representation would factor out the three semilattice equations which \textit{do not} involve sharing and copying (namely,  associativity, commutativity, and idempotence) --- however, a limitation of the current work is that we do not achieve this.  Presently,  we will treat these equations via rewriting as well.

@@

We do not expect internal interfaces to be preserved during rewriting: for example,  when the semilattice equations are considered as rewrites. 
Thus,  we wish a rewrite rule to be a pair of \textit{extended} cospans of e-hypergraphs with matching \textit{external} (but not necessarily internal) interfaces,  as follows. 
\[
    n \xrightarrow{} n' \xrightarrow{} \mathcal{L} \xleftarrow{} m' \xleftarrow{} m
\qquad
    n \xrightarrow{} n'' \xrightarrow{} \mathcal{R} \xleftarrow{} m'' \xleftarrow{} m
\] 
Analogous to Section \ref{sec:combinatorial-semantics}, observing that the following extended cospans express the same data
\[
    0 \xrightarrow{} 0 \xrightarrow{} \mathcal{L} \xleftarrow{} n'' + m'' \xleftarrow{} n + m
\qquad
    0 \xrightarrow{} 0 \xrightarrow{} \mathcal{R} \xleftarrow{} n' + m' \xleftarrow{} n + m
\] allows us to encode rewrite rules as extended cospans of the following form
\[
\mathcal{L} \xleftarrow{} n' + m' \xleftarrow{} n + m \xrightarrow{} n'' + m'' \xrightarrow{} \mathcal{R}
\] which will fit into the DPO formalism. We make this identification throughout,  without further mention . 
%We will futher omit the arrows $0 \to 0 \to \mathcal{L}$.
\begin{remark}
There is a limitation of extended DPOI rewriting as defined in this section: it is unable to express rewriting
for rules whose target contains an edge with no incident vertices.  
While this limitation can be overcome,  we find the technical development to be significantly simpler if we simply avoid consideration of such edges.  
For this reason,  in the remainder of the paper we consider signatures which do not contain $c: 0 \to 0$,  thus restricting possible e-hypergraphs to those without such problematic edges. 
%This is because there is no way to impose both consistency and child relation of such an edge as it has no incident nodes in its context.
%We claim that this is only a minor limitation because in practice such edges (morphisms) with no
%incident nodes correspond to ‘scalars’, if we consider for example a category of vector spaces, and scalars are usually removed by being absorbed by vectors rather than introduced during rewrites.
\end{remark}

@@

Intuitively the above definition means that when the occurence of $\mathcal{L}$ within $\mathcal{G}$ is top-level then $\mathcal{L}$'s external interfaces become the part of $\mathcal{L}^{\bot}$'s external interfaces and when the occurence is nested, those interfaces become the part of $\mathcal{L}^{\bot}$'s strict internal interfaces.
In both cases when removing the occurence of $\mathcal{L}$ from $\mathcal{G}$, $\mathcal{L}$'s strict internal interfaces are removed from strict internal interfaces of $\mathcal{G}$ to form internal interfaces of the corresponding mda-cospans above.
The morphisms are then constructed correspondingly to require that inputs (respectively, outputs) of $\mathcal{R}$ are glued to the outputs (respectively, inputs) of $\mathcal{L}^{\bot}$.

@@

More precisely, when removing the occurence of $\mathcal{L}$ from $\mathcal{G}$, \emph{i.e.}, by computing the pushout complement, the internal interfaces of the resulting e-hypergraph should be modified to \textit{exclude} the vertices corresponding to $\mathcal{L}$'s strictly internal interfaces (since the vertices they map to have been removed). 
Then, when gluing $\mathcal{R}$ into the hole,  the internal interfaces of the resulting e-hypergraph should be modified to \textit{include} the strictly internal interfaces of $\mathcal{R}$ (since new internal interfaces for them to map to have been added).

\begin{definition}[Convex EDPOI rewriting]
\label{def:dpoi-e}
Given an extended span of morphisms 
\[
    \mathcal{G} \xleftarrow{} n' + k' \xleftarrow{} n + k \xrightarrow{} n'' + k'' \xrightarrow{} \mathcal{H}
\]
in $\catname{EHyp(\Sigma)}$, we say $\mathcal{G}$ \textit{rewrites (convexly) to} $\mathcal{H}$ (\textit{with external interface} $n + k$ and \textit{taking internal interface} $n'+k'$ \textit{to} $n'' + k''$) --- denoted by $\mathcal{G} \Rrightarrow \mathcal H$  --- \textit{via a rewrite rule} 
\[
    \mathcal{L} \xleftarrow{} i' + j' \xleftarrow{} i + j \xrightarrow{} i'' + j'' \xrightarrow{} \mathcal{R}
\] 
if there exists an object $\mathcal{C}$ and morphisms which complete the following commutative diagram 
\[
 \scalebox{0.75}{
    \tikzfig{combinatorial_semantics/DPOI_square}
 }
\]
such that the two marked squares are pushouts and the following conditions hold:
    \begin{enumerate}
        \label{dpoi-e:assumptions}
       % \item $m : \mathcal{L} \to \mathcal{G}$ is a convex match;
        %\item $m(L)$ is a down-closed e-hypergraph;
        \item $i + j \to \mathcal{C} \to \mathcal{G}$ is a boundary complement;
	\item the internal interfaces of $\mathcal H$ are such that:
	\[
		n'' = n' \setminus (i' \setminus i) + (i'' \setminus i) \qquad k'' = k' \setminus (j' \setminus j) +  (j'' \setminus j)\\
	\]
	\item the map $f_1 = [g_1,h_1]: n'' \to \mathcal H$ in the diagram above consists of $g_1$ as defined in Definition \ref{def:boundary_new} of boundary complement,  and $h_1$ which is the restiction of the composite $i'' \to \mathcal R \to \mathcal H$ to $i'' \setminus i$,  and similarly for the map $f_2: k'' \to \mathcal H$.  
    \end{enumerate}
% The internal interfaces of the resulting e-hypergraph $\mathcal{H}$ from the cospan
%  \[
%     \mathcal{H} \xleftarrow{} n'' + k'' \xleftarrow{} n + k
% \]
    Given  a set $\mathfrak{R}$ of EDPOI rewrite rules and e-hypergraphs with extended interfaces $\mathcal G$ and $\mathcal H$,  we write $\mathcal G \Rrightarrow_\mathfrak{R} \mathcal H$ if there exists a EDPOI rewrite in $\mathfrak R$ such that via it $\mathcal G$ rewrites to $\mathcal H$,  and we write $\mathcal G \Rrightarrow^*_\mathfrak{R} \mathcal H$ if there exists a sequence of such rewrites. 
\end{definition}
The boundary complement conditions, in particular, prohibits finding a match for $\mathcal L$ in $\mathcal G$,  below.  
\[
	\mathcal{L} = \scalebox{0.8}{\tikzfig{combinatorial_semantics/f_times_g}} \qquad \mathcal{G} = \scalebox{0.6}{\tikzfig{combinatorial_semantics/f_plus_g_inline}}
\]
Together, the conditions ensure the result of a rewrite is a well-typed MDA e-hypergraph.

@@

By construction we have the following corollaries for $<_{\sfrac{X+Y}{\sim}}$ and $\consistency_{\sfrac{X+Y}{\sim}}$.

>>>>

Although equality saturation is already a state-of-the-art algorithmic optimisation technique, there is an opportunity to better understand its mathematical foundations. 
We propose to address this issue via a categorical axiomatisation for e-graphs and their rewriting. 
Considering programs as represented by terms of an arbitrary algebraic theory, we demonstrate a correspondence between e-graphs and (string diagrams for) simple \textit{semilattice-enriched} Cartesian categories. 
The approach we take is guided by the established methodology of creating a correspondence between string diagrams and graph rewriting, as encountered in a large body of research. 
The starting innovation of our approach is the semilattice-enrichment. 
The \textit{join} operator lifts the formalism so that it can now operate with (non-empty) \textit{sets} of terms (string diagrams, morphisms); the rest of the paper is about working out the mathematical implications of introducing this enrichment to string diagrams and their concrete combinatorial representation.

@@

The translation from e-graphs to enriched string diagrams is illustrated informally below,
noting how the typing constraints of the $+$ constructor are satisfied in the image of the translation by discarding in each component all unnecessary inputs.
\[
    \tikzfig{categorical-semantics/egraph-translation-1}
\]
Examples of this translation are given in the second row of Figure \ref{fig:e-graph-example}. In particular, note that the translation of the \textit{cyclic} e-graph (e)
requires the use of a categorical \textit{trace}, generating a cycle. In this paper, we focus on \textit{acyclic} e-graphs and their corresponding categories; but the above example illustrates that the extension to the cyclic case seems to be routine. As it would involve needless additional presentational complexity we leave it as further work.

@@

\subsection{Symmetric Monoidal Theories and PROPs.}

A presentation of an algebraic theory is traditionally given by a \textit{signature} of $n$-ary operations and a set of \textit{equations}---formally,  pairs of terms freely generated over the signature.  We are interested here in symmetric monoidal theories, which are generalisations of algebraic theories where the operations of the signature may have arbitrary \textit{co-arities}.  First,  we generalise the notion of a signature. 
\begin{definition}[Monoidal signature]
A \textit{(monoidal) signature} $\Sigma$ is given by a set of \textit{generators} $c: n \to m$,  with \textit{arity} $n$ and \textit{co-arity} $m$.  %A \textit{signature homomorphism} $h: \Sigma \to \Gamma$ is a function from $\Sigma$ to $\Gamma$ which preserves the (co-)arities of elements. 
\end{definition}

@@

We will use enrichment to axiomatize the e-box structure as the ``join" of two morphisms $f,g: A \to B$ as $f + g: A \to B$. Note, in particular,  that semilattices do not require a unit for $+$. This is important, since, in any commutative monoid enriched category,  if there exists a categorical product, then it must be a categorical biproduct \cite{maclane}---a degeneracy we wish to avoid.  More precisely our hom-sets will be given the structure of a semilattice.  
Generally,  we may define a category \textit{enriched} in any other monoidal category $M$ (\textit{e.g.},  \textbf{SLatt}).  In this case,  hom-sets are generalised to hom-objects of $M$ (\textit{e.g.},  a semilattice) and composition 
\[
	\circ: Hom(B,C) \otimes Hom(A,B) \to Hom(A,C)
\]
is defined as an $M$-morphism.  In other words,  composition must respect the additional structure on hom-sets.  
In our case,  the category we wish to enrich is also monoidal,  so we additionally ask that the monoidal structure also respects this structure.  Technically, this amounts to asking that the tensor product is an enriched functor. 
We omit further details of the general case of enrichment here, and instead work with the following more concrete definition, where $M = \textbf{SLatt}$.

@@

In ~\cite{bonchi_string_2022-1},  hypergraphs \textit{with interfaces} and their DPO rewriting are shown to be sound and complete for SMTs including a Frobenius algebra. 
Hypergraphs generalise graphs by allowing edges to have multiple sources and targets. 
When interpreting string diagrams as hypergraphs,  morphisms are interpreted as edges,  and wires as vertices.  
However,  there is also a need to specify which vertices are to be considered as \textit{input} and \textit{outputs},  corresponding to the dangling wires at the bottom and top of a string diagram. 
This gives rise to the definition of a hypergraph with interfaces.  
Subsequent work \cite{bonchi_string_2022-2} builds on this correspondence to give a restriction of hypergraphs with interfaces (called \textit{monogamous} directed acyclic),  leading to a correspondence between symmetric monoidal string diagrams and the category of such hypergraphs.

@@

The intuitive notion of (hyper-)graph rewriting is  as expected: given a rewrite rule $\mathcal L \rightsquigarrow \mathcal R$ and a graph $\mathcal{G}$,  we expect to find an occurrence of $\mathcal L$ within $\mathcal{G}$,  remove it,  and then plug $\mathcal R$ into the hole to recover the rewritten graph $\mathcal{H}$.  In DPO rewriting,  such a rewrite rule is formalised as a span $\mathcal L \xleftarrow{} \mathcal K \xrightarrow{} \mathcal R$ in some appropriate category of graphs,  where $\mathcal K$ is some invariant that should be maintained during rewriting.
A typical DPO square is depicted in Figure~\ref{fig:dpo_dpoi}~(left).

@@

To express the rewriting of (hyper-)graphs with interfaces,  \textit{i.e.},  the ones that represent morphisms in $\HypI{\Sigma}$, the DPO formalism has been extended to double-pushout rewriting with \textit{interfaces}  (DPOI) as shown in Figure~\ref{fig:dpo_dpoi}~(right).
In the case of $ \catname{Hyp(\Sigma)}$,  we wish rewrite rules to be pairs of discrete cospans of hypergraphs with matching interfaces: $n \xrightarrow{} \mathcal L \xleftarrow{} m$, $n \xrightarrow{} \mathcal R \xleftarrow{} m$.  
By noting that a cospan $0 \xrightarrow{} \mathcal L \xleftarrow{[f,g]} n + m$ encodes the same data as a cospan $n \xrightarrow{f} \mathcal L \xleftarrow{g} m$ (Remark 3.14~\cite{bonchi_string_2022-1}) we can formulate a rewrite rule as a single span $\mathcal L \xleftarrow{} n+m \xrightarrow{} \mathcal R$,  and we can do similarly for the discrete cospans giving the interface to $\mathcal G$ and $\mathcal H$.  
We now make this identification throughout without further mention.  
Thus we reach the following variant of DPO rewriting.
\begin{definition}[DPOI rewriting]\label{def:dpoi}
Given a span of morphisms $\mathcal G \xleftarrow{} n+m \xrightarrow{} \mathcal H$ in $\textbf{Hyp}(\Sigma)$,  we say \textit{$\mathcal G$ rewrites to $\mathcal H$ with interface $n+m$ via rewrite rule $\mathcal L \xleftarrow{} i+j \xrightarrow{} \mathcal R$} if there exists an object $\mathcal C$ and morphisms which complete the following commutative diagram such that the two marked squares are pushouts.
% \[
% \begin{tikzcd}
% \mathcal L \arrow[d, "m"'] & i+j \arrow[d] \arrow[l] \arrow[r] & \mathcal R \arrow[d]   \\
% \mathcal{G}^{\urcorner}     & \mathcal C \arrow[l] \arrow[r]                      & ^{\ulcorner}\mathcal H \\
%                   & n+m \arrow[lu] \arrow[u] \arrow[ru]          &              
% \end{tikzcd}
% \]
\begin{figure}[t!]
    \begin{subfigure}{0.4\linewidth}
    \[
            \scalebox{0.8}{
                \tikzfig{combinatorial_semantics/DPO_square}
            }
    \] 
    % \subcaption{DPO square}
    % \label{fig:dpo}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.4\linewidth}
        \[
    \scalebox{0.8}{
        \tikzfig{combinatorial_semantics/DPOI_square_Hyp}
    }
\]
    % \subcaption{DPOI square}
    % \label{fig:dpoi}
    \end{subfigure}
    \captionsetup{belowskip=-3ex}
    \caption{DPO and DPOI squares}
    \label{fig:dpo_dpoi}
\end{figure}

\end{definition}
We need to restrict the notions of pushout complement and of match in order to preserve monogamy and directed acyclicity.

@@

Hierarchical (hyper-)graphs are a well-studied area of research \cite{plump:hierarchical-graphs, montanari:gs-lambda, palacz:hierarchical-transform, Gaducci:hierarchical-graphs, Ghica:hierarchical}.  In this section,  we follow analogous steps to those used to achieve a combinatorial representation of $\textbf{PROP}(\Sigma)$ by $\mathbf{MHypI}(\Sigma)$,  but in the semilattice enriched setting.  First we define an \textit{e-hypergraph},  which supports a hierarchical notion of ``e-box'' with distinct components. Then,  we extend this notion with interfaces via a generalisation of the cospan construction.  Finally,  we restrict the morphisms under consideration to those satisfying the MDA condition,  as in the previous section.  An example diagram representing e-hypergraphs with interfaces can be found in Figure \ref{fig:A+B}.

@@

When modelling enriched string diagrams using e-hypergraphs,  we require to keep track of more than simply the ultimate external inputs and outputs of the diagram: each internal e-box has its own inputs and outputs,  which we must also keep track of.  Thus,  we introduce an \textit{extended} cospan construction.  Since the $\catname{EHyp(\Sigma)}$ does not have pushouts in general,  we skip straight to considering a category of extended cospans of discrete e-hypergraphs (\textit{i.e.},  discrete hypergraphs considered as e-hypergraphs in the obvious way),  for which pushouts do exist.

We use the notation $n \setminus m$ to denote the discrete e-hypergraph with $n - m$ vertices,  in particular with the vertices of $m$ removed from $n$ when vertices of $m$ is a sub-e-hypergraph of $n$.
\begin{definition}[Category of e-hypergraphs with interfaces]
    The category of \textit{e-hypergraphs with extended interfaces} $\Ecospans$ has discrete e-hypergraphs as objects,  with hom-sets $\Ecospans(n,m)$ consisting of isomorphism classes of \textit{extended cospans}, defined as follows:  
    \[
    n \xrightarrow{f_{ext}} n' \xrightarrow{f_{int}} \mathcal{G} \xleftarrow{g_{int}} m' \xleftarrow{g_{ext}} m
    \]
    where $\mathcal{G}$ is an e-hypergraph,  and $n, n', m, m'$ are discrete e-hypergraphs,  $f_{ext},f_{int},g_{ext},g_{int}$ are monomorphisms in $\catname{EHyp(\Sigma)}$,  and the image of $f_{ext};f_{int}$ and of $g_{ext};g_{int}$ consist exclusively of top-level vertices,  and that that vertices in the strictly internal interface (defined below) are not top-level.  We will sometimes write $\mathcal G$ to denote the extended cospan,  where it is clear from context $\mathcal G$ is equipped with extended interfaces.

@@

Unlike in the previous section,  in order to give an equivalence with $\textbf{PROP}^+(\Sigma)$,  we must first develop the theory of DPOI rewriting for e-hypergraphs.  This is because the equations for semilattice enrichment are not intended to be subsumed by the e-hypergraph representation.  Instead,  we will treat them via rewriting,  as we do with equations of the signature.  This is because the semilattice equations involve sharing and copying, operations on string diagrams which are not usually quotiented away, cf. string diagrams for Cartesian categories.  An ideal combinatorial representation would factor out the three semilattice equations which \textit{do not} involve sharing and copying (namely,  associativity, commutativity, and idempotence) --- however, a limitation of the current work is that we do not achieve this.  Presently,  we will treat these equations via rewriting as well.

@@

We do not expect internal interfaces to be preserved during rewriting: for example,  when the semilattice equations are considered as rewrites. 
Thus,  we wish a rewrite rule to be a pair of \textit{extended} cospans of e-hypergraphs with matching \textit{external} (but not necessarily internal) interfaces,  as follows. 
\[
    n \xrightarrow{} n' \xrightarrow{} \mathcal{L} \xleftarrow{} m' \xleftarrow{} m
\qquad
    n \xrightarrow{} n'' \xrightarrow{} \mathcal{R} \xleftarrow{} m'' \xleftarrow{} m
\] 
Analogous to Section \ref{sec:combinatorial-semantics}, observing that the following extended cospans express the same data
\[
    0 \xrightarrow{} 0 \xrightarrow{} \mathcal{L} \xleftarrow{} n'' + m'' \xleftarrow{} n + m
\qquad
    0 \xrightarrow{} 0 \xrightarrow{} \mathcal{R} \xleftarrow{} n' + m' \xleftarrow{} n + m
\] allows us to encode rewrite rules as extended cospans of the following form
\[
\mathcal{L} \xleftarrow{} n' + m' \xleftarrow{} n + m \xrightarrow{} n'' + m'' \xrightarrow{} \mathcal{R}
\] which will fit into the DPO formalism. We make this identification throughout,  without further mention. 
%We will futher omit the arrows $0 \to 0 \to \mathcal{L}$.
\begin{remark}
There is a limitation of extended DPOI rewriting as defined in this section: it is unable to express rewriting
for rules whose target contains an edge with no incident vertices.  
While this limitation can be overcome,  we find the technical development to be significantly simpler if we simply avoid consideration of such edges.  
For this reason,  in the remainder of the paper we consider signatures which do not contain $c: 0 \to 0$,  thus restricting possible e-hypergraphs to those without such problematic edges. 
%This is because there is no way to impose both consistency and child relation of such an edge as it has no incident nodes in its context.
%We claim that this is only a minor limitation because in practice such edges (morphisms) with no
%incident nodes correspond to ‘scalars’, if we consider for example a category of vector spaces, and scalars are usually removed by being absorbed by vectors rather than introduced during rewrites.
\end{remark}

@@

Intuitively the above definition means that when the occurrence of $\mathcal{L}$ within $\mathcal{G}$ is top-level then $\mathcal{L}$'s external interfaces become the part of $\mathcal{L}^{\bot}$'s external interfaces and when the occurrence is nested, those interfaces become the part of $\mathcal{L}^{\bot}$'s strict internal interfaces.
In both cases when removing the occurrence of $\mathcal{L}$ from $\mathcal{G}$, $\mathcal{L}$'s strict internal interfaces are removed from strict internal interfaces of $\mathcal{G}$ to form internal interfaces of the corresponding mda-cospans above.
The morphisms are then constructed correspondingly to require that inputs (respectively, outputs) of $\mathcal{R}$ are glued to the outputs (respectively, inputs) of $\mathcal{L}^{\bot}$.

@@

More precisely, when removing the occurrence of $\mathcal{L}$ from $\mathcal{G}$, \emph{i.e.}, by computing the pushout complement, the internal interfaces of the resulting e-hypergraph should be modified to \textit{exclude} the vertices corresponding to $\mathcal{L}$'s strictly internal interfaces (since the vertices they map to have been removed). 
Then, when gluing $\mathcal{R}$ into the hole,  the internal interfaces of the resulting e-hypergraph should be modified to \textit{include} the strictly internal interfaces of $\mathcal{R}$ (since new internal interfaces for them to map to have been added).

\begin{definition}[Convex EDPOI rewriting]
\label{def:dpoi-e}
Given an extended span of morphisms 
\[
    \mathcal{G} \xleftarrow{} n' + k' \xleftarrow{} n + k \xrightarrow{} n'' + k'' \xrightarrow{} \mathcal{H}
\]
in $\catname{EHyp(\Sigma)}$, we say $\mathcal{G}$ \textit{rewrites (convexly) to} $\mathcal{H}$ (\textit{with external interface} $n + k$ and \textit{taking internal interface} $n'+k'$ \textit{to} $n'' + k''$) --- denoted by $\mathcal{G} \Rrightarrow \mathcal H$  --- \textit{via a rewrite rule} 
\[
    \mathcal{L} \xleftarrow{} i' + j' \xleftarrow{} i + j \xrightarrow{} i'' + j'' \xrightarrow{} \mathcal{R}
\] 
if there exists an object $\mathcal{C}$ and morphisms which complete the following commutative diagram 
\[
 \scalebox{0.75}{
    \tikzfig{combinatorial_semantics/DPOI_square}
 }
\]
such that the two marked squares are pushouts and the following conditions hold:
    \begin{enumerate}
        \label{dpoi-e:assumptions}
       % \item $m : \mathcal{L} \to \mathcal{G}$ is a convex match;
        %\item $m(L)$ is a down-closed e-hypergraph;
        \item $i + j \to \mathcal{C} \to \mathcal{G}$ is a boundary complement;
	\item the internal interfaces of $\mathcal H$ are such that:
	\[
		n'' = n' \setminus (i' \setminus i) + (i'' \setminus i) \qquad k'' = k' \setminus (j' \setminus j) +  (j'' \setminus j)\\
	\]
	\item the map $f_1 = [g_1,h_1]: n'' \to \mathcal H$ in the diagram above consists of $g_1$ as defined in Definition \ref{def:boundary_new} of boundary complement,  and $h_1$ which is the restriction of the composite $i'' \to \mathcal R \to \mathcal H$ to $i'' \setminus i$,  and similarly for the map $f_2: k'' \to \mathcal H$.  
    \end{enumerate}
% The internal interfaces of the resulting e-hypergraph $\mathcal{H}$ from the cospan
%  \[
%     \mathcal{H} \xleftarrow{} n'' + k'' \xleftarrow{} n + k
% \]
    Given  a set $\mathfrak{R}$ of EDPOI rewrite rules and e-hypergraphs with extended interfaces $\mathcal G$ and $\mathcal H$,  we write $\mathcal G \Rrightarrow_\mathfrak{R} \mathcal H$ if there exists a EDPOI rewrite in $\mathfrak R$ such that via it $\mathcal G$ rewrites to $\mathcal H$,  and we write $\mathcal G \Rrightarrow^*_\mathfrak{R} \mathcal H$ if there exists a sequence of such rewrites. 
\end{definition}
The boundary complement conditions, in particular, prohibits finding a match for $\mathcal L$ in $\mathcal G$,  below.  
\[
	\mathcal{L} = \scalebox{0.8}{\tikzfig{combinatorial_semantics/f_times_g}} \qquad \mathcal{G} = \scalebox{0.6}{\tikzfig{combinatorial_semantics/f_plus_g_inline}}
\]
Together, the conditions ensure the result of a rewrite is a well-typed MDA e-hypergraph.

@@

% By construction we have the following corollaries for $<_{\sfrac{X+Y}{\sim}}$ and $\consistency_{\sfrac{X+Y}{\sim}}$.
